Supplemental material for: Timing and magnitude of climate driven range shifts in transboundary fish stocks challenge their management

### 2021-08-24 Scripts/Manuscript.Rmd
updates for submission to gcb

---
title: "Timing and magnitude of climate driven range shifts in transboundary fish stocks challenge their management"
author: "REVISED MANUSCRIPT. DO NOT SHARE."
date: "Affiliations: ^1^Institute for the Oceans and Fisheries, University of British Columbia, Vancouver, Canada. ^2^Climate and Environmental Physics, Physics Institute, University of Bern. ^3^Oeschger Centre for Climate Change Research, University of Bern, Switzerland. ^4^School of Environmental Sciences, University of Liverpool. ^5^Stanford Center for Ocean Solutions, Stanford University, Stanford, United States"
subtitle: "Juliano Palacios-Abrantes*^1,2^, Thomas L. Frölicher^3,4^, Gabriel Reygondeau^1^, U. Rashid Sumaila^1^, Alessandro Tagliabue^5^, Colette C.C. Wabnitz^1,6^, and William W.L. Cheung^1^"
# output: bookdown::pdf_document2
output: 
  bookdown::word_document2:
    number_section: no
reference_docx:  ../References/sci_template.docx
bibliography: ../References/ToE_Ref.bib
csl: ../References/pnas.csl
header-includes:
  - \usepackage{setspace}
  - \usepackage{caption}
  - \usepackage{float} # Use the 'float' package
  - \captionsetup[figure]{name = Fig}
---

@@

**Email**: [j.palacio\@oceans.ubc.ca](mailto:j.palacios@oceans.ubc.ca){.email}

**Author Contributions:**  
Conceptualization: JPA, WWLC, TF, AT
Data curation; TF; WWLC; GR
Formal Analysis: JPA
Methodology: JPA, WWLC, TF, AT, GR, CCCW
Visualization: JPA
Writing – original draft: JPA; WWLC; CCCW; TF
Writing – review & editing: JPA; WWLC; CCCW; TF; AT: URS; GR

**Competing Interest Statement:** Authors declare that they have no competing interests.

Classification: Biological Science, Environmental Sciences

This PDF file includes:
Main Text
Figures 1 to 5

@@

# Abstract {-}

<!-- An opening sentence that states the question/problem addressed by the research AND Enough background content to give context to the study AND A brief statement of primary results AND A short concluding sentence. -->

Climate change is shifting the distribution of shared fish stocks between neighboring countries’ Exclusive Economic Zones (EEZ) and the high seas. The timescale of these transboundary shifts determines how climate change will affect international fisheries governance. Coupling a large ensemble simulation of an Earth system model to a species distribution model, we show that by 2030 23% of transboundary stocks would have shifted and 78% of the world will experience at least one shifting stock. By 2100, 81% of EEZ waters will see at least one shifting stock with a total of 45% of stocks shifting globally, under a high emissions scenario. Importantly, many countries that are highly dependent on fisheries for income, food and nutrition security, as well as livelihoods emerge as hotspots for transboundary shifts. Existing fisheries agreements need to be assessed for their capacity in addressing transboundary shifts, and strengthened where necessary to limit conflict over these fish stocks.

# Significant statement {-}

Marine fish stocks are shifting their distribution in response to climate change. This is of particular importance for the management of stocks shared between neighboring Exclusive Economic Zones, also known as transboundary stocks. Here, we find that imminent climate change-induced shifts in transboundary fish stocks are challenging international management and that urging fisheries agreements to be assessed for their capacity in addressing transboundary shifts, and strengthened where necessary to limit conflict over these fish stocks.

# Main text {-}

@@

```{r toeFigure1N, eval = T,  echo = F, fig.width=14, fig.height=10, fig.cap =  "Schematic diagram of the transboundary index (TI) to determine the time of emergence of transboundary stocks from one Exclusive Economic Zone (EEZ) to another. The index is based on the distance between the distributional centroid of the transboundary stock and the geographic centroid of the neighboring EEZ sharing the stock (Top panel). Time of emergence is defined as the first year when the TI overshoots historical values (Bottom panel). Only showing a subset of ensemble members for clarity."}

@@

We coupled a mechanistic population dynamic model with a ten-ensemble member simulation of an Earth System Model to project the future global distribution of transboundary stocks under climate change (Methods).

@@

# Results {-}

### Identifying range shifts in the shared distribution of transboundary stocks {-}

### Shifting intensity of transboundary stocks {-}

# Discussion {-}

### High present-day climate risk for transboundary fisheries {-}

### Climate risk hotspots for transboundary fisheries management {-}

### Potential drivers of shift and managing an uncertain future {-}

# Materials and Methods {-}

@@

The analyses detailed herein are based on 633 exploited marine transboundary species that account for 80% of the catch taken from the world’s EEZs between 2005 and 2014 [@PalaciosAbrantes:2020tv]. When a species was shared by a pair of neighboring EEZs it was considered a transboundary stock [@PalaciosAbrantes:2020tv; @Teh:2015gd], resulting in a total of 9,132 transboundary stocks. Using the transboundary species Atlantic cod (*Gadus morhua*) as an example and this study’s definition of transboundary stock, the United States and Canada share a stock of Atlantic cod, and Canada and Greenland share a separate stock of Atlantic cod, but the United States and Greenland do not as they do not have juxtaposing EEZs. We defined the boundaries of the world’s EEZs using the Sea Around Us spatial division (updated 1 July 2015; http://www.seaaroundus.org), noting that it subdivides the EEZs of 198 coastal states into 280 regions (Fig. \@ref(fig:toeFigureS5)) including island territories. We determined the intersections between polygons using the R package sf [@PackagesfSimpleF:2018vp]. Each EEZ was categorized by geopolitical region according to the United Nations (https://population.un.org/wpp/DefinitionOfRegions/) and biome [@Reygondeau:2019uh]. The habitat preference of each species was determined following FishBase (http://www.fishbase.org) for fish species and SeaLifeBase (http://www.fishbase.org) for invertebrates (Table S\@ref(tab:toeTableS1)). For each stock and EEZ, we used Sea Around Us data to estimate catch and fishing revenue from fishing activities. We report both average catch and revenue for the last available decade (2005-2014) [@Zeller2016; @Tai2017; @Sumaila:2015uc].

@@

# Acknowledgments {-}
We would like to acknowledge inputs from T. Clarke and L. Kwong on previous iterations of the manuscript and figures. A.T. acknowledges  funding from the Horizon 2020 Framework Programme (724289). C.C.C.W acknowledges funding from the David and Lucile Packard Foundation (2019-68336), the Gordon and Betty Moore Foundation (GBMF5668.02), and the Walton Family Foundation (2018-1371). J.P.A and W.W.L.C. acknowledge funding from the Social Sciences and Humanities Research Council of Canada, Partnership Grant and the Natural Sciences and Engineering Research Council of Canada, Partnership Grant. T.F. acknowledges funding from the Horizon 2020 Framework Programme (820989), the Swiss National Science Foundation (PP00P2_170687), and the Swiss National Supercomputing Centre. U.R.S. acknowledges funding from the Social Sciences and Humanities Research Council of Canada, Partnership Grant.

# References {-}

# Figures and Tables

```{r toeFigure1N, eval = T,  echo = F, fig.width=14, fig.height=10, fig.cap =  "Schematic diagram of the transboundary index (TI) to determine the time of emergence of transboundary stocks from one Exclusive Economic Zone (EEZ) to another. The index is based on the distance between the distributional centroid of the transboundary stock and the geographic centroid of the neighboring EEZ sharing the stock (Top panel). Time of emergence is defined as the first year when the TI overshoots historical values (Bottom panel). Only showing a subset of ensemble members for clarity."}

@@

@@

>>>>

---
title: "Timing and magnitude of climate driven range shifts in transboundary fish stocks challenge their management"
date: "Affiliations: ^1^Institute for the Oceans and Fisheries, University of British Columbia, Vancouver, Canada.^2^Center for Limnology, University of Wisconsin, Madison. ^3^Climate and Environmental Physics, Physics Institute, University of Bern. ^4^Oeschger Centre for Climate Change Research, University of Bern, Switzerland. ^5^School of Environmental Sciences, University of Liverpool. ^6^Stanford Center for Ocean Solutions, Stanford University, Stanford, United States"
subtitle: "Juliano Palacios-Abrantes*^1,2^, Thomas L. Frölicher^3,4^, Gabriel Reygondeau^1^, U. Rashid Sumaila^1^, Alessandro Tagliabue^5^, Colette C.C. Wabnitz^1,6^, and William W.L. Cheung^1^"
output: 
  bookdown::pdf_document2:
    number_section: no
    toc: no
    latex_engine: xelatex
reference_docx:  ../References/sci_template.docx
bibliography: ../References/ToE_Ref.bib
csl: ../References/gcb.csl
header-includes:
  - \usepackage{setspace}
  - \usepackage{caption}
  - \usepackage[left]{lineno} # for numbers in lines
  - \usepackage{float} # Use the 'float' package
  - \captionsetup[figure]{name = Fig}
  - \floatplacement{figure}{H}
---

<!-- Running title: Imminent shifts of transboundary fish stocks -->

@@

<!-- **Email**: [j.palacio\@oceans.ubc.ca](mailto:j.palacios@oceans.ubc.ca){.email} -->

<!-- **Author Contributions:**   -->
<!-- Conceptualization: JPA, WWLC, TF, AT -->
<!-- Data curation; TF; WWLC; GR -->
<!-- Formal Analysis: JPA -->
<!-- Methodology: JPA, WWLC, TF, AT, GR, CCCW -->
<!-- Visualization: JPA -->
<!-- Writing – original draft: JPA; WWLC; CCCW; TF -->
<!-- Writing – review & editing: JPA; WWLC; CCCW; TF; AT: URS; GR -->

<!-- **Competing Interest Statement:** Authors declare that they have no competing interests. -->

<!-- Thomas Frölicher -->
<!-- Horizon 2020 Framework Programme 820989 -->
<!-- Swiss National Science Foundation PP00P2_170687 -->
<!-- Swiss National Supercomputing Centre -->

<!-- Alessandro Tagliabue -->
<!-- Horizon 2020 Framework Programme  724289 -->

<!-- Juliano Palacios-Abrantes -->
<!-- Natural Sciences and Engineering Research Council of Canada  Discovery Grant -->
<!-- Social Sciences and Humanities Research Council of Canada Partnership Grant -->

<!-- William Cheung -->
<!-- Natural Sciences and Engineering Research Council of Canada - Discovery Grant -->
<!-- Social Sciences and Humanities Research Council of Canada - Partnership Grant -->

<!-- U. Rashid Sumaila -->
<!-- Social Sciences and Humanities Research Council of Canada --> <!-- Partnership Grant -->

<!-- Colette C.C. Wabnitz -->
<!-- David and Lucile Packard Foundation 2019-68336 -->
<!-- Gordon and Betty Moore Foundation GBMF5668.02 -->
<!-- Walton Family Foundation 2018-1371 -->

\linenumbers

# Abstract

Climate change is shifting the distribution of shared fish stocks between neighboring countries’ Exclusive Economic Zones (EEZ) and the high seas. The timescale of these transboundary shifts determines how climate change will affect international fisheries governance. Coupling a large ensemble simulation of an Earth system model to a species distribution model, we show that by 2030, 23% of transboundary stocks would have shifted and 78% of the world EEZs will experience at least one shifting stock under a high emission climate change scenario. By the end of this century, 81% of EEZ waters will see at least one shifting stock with a total of 45% of stocks shifting globally, under a high emissions scenario. Importantly, many countries that are highly dependent on fisheries for income, food and nutrition security, as well as livelihoods emerge as hotspots for transboundary shifts showing early, and sometimes past shifts. Existing fisheries agreements need to be assessed for their capacity in addressing transboundary shifts, and strengthened where necessary to limit conflict over these fish stocks while new agreements are urged to considere this problematic in order to be resilient to global change.

<!-- # Significant statement {-} -->

<!-- Marine fish stocks are shifting their distribution in response to climate change. This is of particular importance for the management of stocks shared between neighboring Exclusive Economic Zones, also known as transboundary stocks. Here, we find that imminent climate change-induced shifts in transboundary fish stocks are challenging international management and that urging fisheries agreements to be assessed for their capacity in addressing transboundary shifts, and strengthened where necessary to limit conflict over these fish stocks. -->

@@

# Introduction

@@

```{r toeFigure1N, eval = T,  echo = F, fig.width=14, fig.height=10, fig.pos="H", fig.cap =  "Schematic diagram of the transboundary index (TI) to determine the time of emergence of transboundary stocks from one Exclusive Economic Zone (EEZ) to another. The index is based on the distance between the distributional centroid of the transboundary stock and the geographic centroid of the neighboring EEZ sharing the stock (Top panel). Time of emergence is defined as the first year when the TI overshoots historical values (Bottom panel). Only showing a subset of ensemble members for clarity."}

@@

<!-- We coupled a mechanistic population dynamic model with a ten-ensemble member simulation of an Earth System Model to project the future global distribution of transboundary stocks under climate change (Methods).  -->

@@

# Materials and Methods

@@

The analyses detailed herein are based on 633 exploited marine transboundary species that account for 80% of the catch taken from the world’s EEZs between 2005 and 2014 [@PalaciosAbrantes:2020tv]. When a species was shared by a pair of neighboring EEZs it was considered a transboundary stock [@PalaciosAbrantes:2020tv; @Teh:2015gd], resulting in a total of 9,132 transboundary stocks. Using the transboundary species Atlantic cod (*Gadus morhua*) as an example and this study’s definition of transboundary stock, the United States and Canada share a stock of Atlantic cod, and Canada and Greenland share a separate stock of Atlantic cod, but the United States and Greenland do not as they do not have juxtaposing EEZs. We defined the boundaries of the world’s EEZs using the Sea Around Us spatial division (updated 1 July 2015; http://www.seaaroundus.org), noting that it subdivides the EEZs of 198 coastal states into 280 regions (Fig. \@ref(fig:toeFigureS5)) including island territories. We determined the intersections between polygons using the R package sf [@PackagesfSimpleF:2018vp]. Each EEZ was categorized by geopolitical region according to the United Nations (https://population.un.org/wpp/DefinitionOfRegions/) and biome [@Reygondeau:2019uh]. The habitat preference of each species was determined following FishBase (http://www.fishbase.org) for fish species and SeaLifeBase (http://www.fishbase.org) for invertebrates (Table \@ref(tab:toeTableS1)). For each stock and EEZ, we used Sea Around Us data to estimate catch and fishing revenue from fishing activities. We report both average catch and revenue for the last available decade (2005-2014) [@Zeller2016; @Tai2017; @Sumaila:2015uc].

@@

# Results

## Identifying range shifts in the shared distribution of transboundary stocks {-}

@@

## Shifting intensity of transboundary stocks

@@

# Discussion

## High present-day climate risk for transboundary fisheries

## Climate risk hotspots for transboundary fisheries management

## Potential drivers of shift and managing an uncertain future

<!-- # Acknowledgments {-} -->
<!-- We would like to acknowledge inputs from T. Clarke and L. Kwong on previous iterations of the manuscript and figures. A.T. acknowledges  funding from the Horizon 2020 Framework Programme (724289). C.C.C.W acknowledges funding from the David and Lucile Packard Foundation (2019-68336), the Gordon and Betty Moore Foundation (GBMF5668.02), and the Walton Family Foundation (2018-1371). J.P.A and W.W.L.C. acknowledge funding from the Social Sciences and Humanities Research Council of Canada, Partnership Grant and the Natural Sciences and Engineering Research Council of Canada, Partnership Grant. T.F. acknowledges funding from the Horizon 2020 Framework Programme (820989), the Swiss National Science Foundation (PP00P2_170687), and the Swiss National Supercomputing Centre. U.R.S. acknowledges funding from the Social Sciences and Humanities Research Council of Canada, Partnership Grant. -->

# References

<!-- # Supplemental material -->

\renewcommand{\thefigure}{S\arabic{figure}}
\setcounter{figure}{0}
\renewcommand{\thetable}{S\arabic{table}}
\setcounter{table}{0}

```{r supplements, child = 'Supplements.Rmd', eval = T}
```

### 2021-08-24 Scripts/Supplements.Rmd
updates for submission to gcb

---
title: "Timing and magnitude of climate driven range shifts in transboundary fish stocks challenge their management"
author: 'Juliano Palacios-Abrantes, Thomas L. Frölicher, Gabriel Reygondeau, U. Rashid Sumaila, Alessandro Tagliabue, Colette C.C. Wabnitz, and William W.L. Cheung'
output:
  # bookdown::word_document2
    bookdown::pdf_document2:
    toc : no
    keep_tex: no
    number_sections: yes
header-includes:
- \usepackage{float} #use the 'float' package
- \floatplacement{figure}{H} #make every figure with caption = h
- \floatplacement{table}{H}
- \usepackage{caption}
- \captionsetup[figure]{name = Figure S}
- \captionsetup[table]{name = Table S}
---

@@

# Supplemental Figures

S1

```{r toeFigureS2, eval = T,  echo = F, fig.width=12, fig.height=7, fig.cap ="Year of change in the shared distribution of 5,745 transboundary stocks using a one standard deviation threshold. A) Land polygons show the contribution of shifting stocks to a country or territory’s total fishing revenue from transboundary stocks. Exclusive Economic Zone polygons display average year of range shifts within them. EEZs with no distributional shift between 2006 and 2100 are represented in aqua color. B) Year of shared distribution shifts by UN sub-regions. Points color coded by the median year of range shift. Horizontal dashed line represents the year 2030 – the Sustainable Development Agenda deadline. N = North, S = South, W = West and E = East. Ltn. Ame. and the Car. = Latin America and the Caribbean. Aus and New Z. = Australia and New Zealand."}

S2

```{r toeFigureS8, eval = T,  echo = F, fig.width=6, fig.height=6, fig.cap ="Average distribution of transboundary stock’s year of change in their shared distribution. Left panel is the average year across stocks and right panel the standard deviation. Yellow represents a one standard deviation distance from the historical average and purple represents a two standard deviation distance from the historical average."}

@@

S3

```{r toeFigureS1, eval = T,  echo = F, fig.width=9, fig.height=7,fig.pos= "H", fig.cap =  "Comparison of shifts in the shared distribution of transboundary stocks by species’ habitat association. Classification is based on the habitat preference information obtained from FishBase and SeaLifeBase (see Table S1). The category “Other” consists of species that have no “habitat preference” classification. The number of stocks included in this analysis for each habitat type is noted in parentheses. Dashed whiskers represent the 1.5* interquartile range. Box represents interquartile range as distance between first and third quartiles. Solid line represents the median time of emergence per habitat preference."}

S4

S5

```{r toeFigureS4, eval = T,  echo = F, fig.width=14, fig.height=7, fig.cap ="Changes in stock proportion of neighboring Exclusive Economic Zones by 2050 (2041-2060) relative to 1951-2005. Lines represent the average change in transboundary stock share ratio with arrows going from EEZ with decreasing stock share (point) to those gaining shares (arrowhead). Land polygons represent the percentage of stocks that are projected to change their stock share ratio beyond the identified threat point with higher gains identified in warmer colors. Panel B highlights changes for the Caribbean and Northwestern Europe."}

S6

@@

S7

```{r toeFigureS5, eval = T,  echo = F, fig.width=4, fig.height=4, fig.cap =  "World Exclusive Economic Zones used in this study as divided by the Sea Around Us and their estimated centroids (black points)."}

@@

_**Tables**_

@@

    "Living and feeding in the open sea; associated with the surface or middle depths of a body of water; free swimming in the seas, oceans or open waters; not in association with the bottom. Many pelagic fish feed on plankton. In FishBase, referring to surface or mid water from 0 to 200 m depth."
  )
) %>% 
  write.csv(.,"habitat_preference.csv")
 kable(
        booktabs = T,
        caption = "Definition of each habitat preference according to FishBase"
  ) #%>% 
  # kable_styling(full_width = F,
                # position = "center")

@@

```{r toeTableS3, eval = F, echo = F}

read.csv("~/GitHub/TransEmergence/Tables/kw_habitat.csv") %>% 
  select(-statistic,
         "Sub region comparrison" = comparrison,
         "Significan level" = signif.level,
         "Observation difference" = dif.com.obs.dif,
         "Critical difference" = dif.com.critical.dif,
          "K-W Difference" = dif.com.difference) %>% 
  kable(
        booktabs = T,
        caption = "Statistical results for multiple comparison test after Kruskal-Wallis on time of emergence across United Nations sub regions"
  ) #%>% 
  # kable_styling(full_width = F,
                # position = "center")

@@

```{r toeTableS2, eval = F, echo = F}

read.csv("~/GitHub/TransEmergence/Tables/kw_regional.csv") %>% 
  select(-statistic,
         "habitat association comparrison" = comparrison,
         "Significan level" = signif.level,
         "Observation difference" = dif.com.obs.dif,
         "Critical difference" = dif.com.critical.dif,
          "K-W Difference" = dif.com.difference) %>% 
  kable(
        booktabs = T,
        caption = "Statistical results for multiple comparison test after Kruskal-Wallis on time of emergence across United Nations sub regions"
  ) #%>% 
  # kable_styling(full_width = F,
                # position = "center")

@@

>>>>

---
title: "Supplemental material for: Timing and magnitude of climate driven range shifts in transboundary fish stocks challenge their management"
date: 'Juliano Palacios-Abrantes, Thomas L. Frölicher, Gabriel Reygondeau, U. Rashid Sumaila, Alessandro Tagliabue, Colette C.C. Wabnitz, and William W.L. Cheung'
output: 
  bookdown::pdf_document2:
    number_section: no
    toc: no
    latex_engine: xelatex
reference_docx:  ../References/sci_template.docx
header-includes:
- \usepackage{float} #use the 'float' package
- \floatplacement{figure}{H} #make every figure with caption = h
- \floatplacement{table}{H}
- \usepackage{caption}
- \captionsetup[figure]{name = Figure S}
- \captionsetup[table]{name = Table S}
- \usepackage{lscape}
---

@@

# Figures

```{r toeFigureS5, eval = T,  echo = F, fig.width = 10, fig.height= 8, fig.pos="H", fig.cap ="World Exclusive Economic Zones used in this study as divided by the Sea Around Us and their estimated centroids (black points)."}

@@

@@

```{r toeFigureS2, eval = T,  echo = F, fig.width=13, fig.height=13, fig.cap ="Year of change in the shared distribution of 5,745 transboundary stocks using a one standard deviation threshold. A) Land polygons show the contribution of shifting stocks to a country or territory’s total fishing revenue from transboundary stocks. Exclusive Economic Zone polygons display average year of range shifts within them. EEZs with no distributional shift between 2006 and 2100 are represented in aqua color. B) Year of shared distribution shifts by UN sub-regions. Points color coded by the median year of range shift. Horizontal dashed line represents the year 2030 – the Sustainable Development Agenda deadline. N = North, S = South, W = West and E = East. Ltn. Ame. and the Car. = Latin America and the Caribbean. Aus and New Z. = Australia and New Zealand."}

```{r toeFigureS8, eval = T,  echo = F, fig.width=4, fig.height=4, fig.cap ="Average distribution of transboundary stock’s year of change in their shared distribution. Left panel is the average year across stocks and right panel the standard deviation. Yellow represents a two standard deviation distance from the historical average and purple represents a one standard deviation distance from the historical average."}

```{r toeFigureS1, eval = T,  echo = F, fig.width=9, fig.height=6,fig.pos= "H", fig.cap =  "Comparison of shifts in the shared distribution of transboundary stocks by species’ habitat association. Classification is based on the habitat preference information obtained from FishBase and SeaLifeBase (see Table S1). The category “Other” consists of species that have no “habitat preference” classification. The number of stocks included in this analysis for each habitat type is noted in parentheses. Dashed whiskers represent the 1.5* interquartile range. Box represents interquartile range as distance between first and third quartiles. Solid line represents the median time of emergence per habitat preference."}

```{r toeFigureS4, eval = T,  echo = F, fig.width=14, fig.height=14, fig.cap ="Changes in stock proportion of neighboring Exclusive Economic Zones by 2050 (2041-2060) relative to 1951-2005. Lines represent the average change in transboundary stock share ratio with arrows going from EEZ with decreasing stock share (point) to those gaining shares (arrowhead). Land polygons represent the percentage of stocks that are projected to change their stock share ratio beyond the identified threat point with higher gains identified in warmer colors. Panel B highlights changes for the Caribbean and Northwestern Europe."}

@@

# Tables

@@

    "Living and feeding in the open sea; associated with the surface or middle depths of a body of water; free swimming in the seas, oceans or open waters; not in association with the bottom. Many pelagic fish feed on plankton. In FishBase, referring to surface or mid water from 0 to 200 m depth."
  )
) %>% 
  # write.csv(.,"habitat_preference.csv")
 kable(
        booktabs = T,
        # longtable = T,
        caption = "Definition of each habitat preference according to FishBase"
  ) %>%
  kable_styling(full_width = T,
                position = "center")

@@

```{r toeTableS3, eval = T, echo = F}

read.csv("../Tables/kw_habitat.csv") %>% 
  select(-statistic,
         "Sub region comparrison" = comparrison,
         "Significan level" = signif.level,
         "Observation difference" = dif.com.obs.dif,
         "Critical difference" = dif.com.critical.dif,
          "K-W Difference" = dif.com.difference) %>% 
  kable(
        booktabs = T,
        longtable = T,
        caption = "Statistical results for multiple comparison test after Kruskal-Wallis on time of emergence across United Nations sub regions"
  ) %>% 
  kable_styling(#full_width = T,
                position = "center")

@@

\begin{landscape}

```{r toeTableS2, eval = T, echo = F}

read.csv("../Tables/kw_regional.csv") %>% 
  select(-statistic,
         "habitat association comparrison" = comparrison,
         "Significan level" = signif.level,
         "Observation difference" = dif.com.obs.dif,
         "Critical difference" = dif.com.critical.dif,
          "K-W Difference" = dif.com.difference) %>% 
  kable(
        booktabs = T,
        longtable = T,
        caption = "Statistical results for multiple comparison test after Kruskal-Wallis on time of emergence across United Nations sub regions"
  ) %>% 
  kable_styling(#full_width = T,
                # position = "center",
                latex_options = c("repeat_header")
                )

@@

\end{landscape}
